---
title: "Getting Started with atspR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with atspR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
library(atspR)
```

## Overview

`atspR` is an R package that provides an automated, modular preprocessing
pipeline for time-series data covering the full journey from raw sensor data
to cross-validated, model-ready train and test sets.

The package is designed for students and domain practitioners (agronomists,
environmental scientists, irrigation engineers) who work primarily in R and
need a reliable, leakage-free preprocessing workflow without having to manage
multiple packages or worry about the correct sequencing of steps.

### Why atspR?

| Problem | How atspR solves it |
|---|---|
| Scalers applied before splitting → **data leakage** | Scaler fitted on **TRAIN only**, applied to both sets |
| Random split invalidates time-series models | **Temporal split** — train always precedes test chronologically |
| Silent NA removal with no record | Full report of what was dropped or imputed, and why |
| Multi-package workflow, easy to sequence incorrectly | Single function `ts_preprocess()` runs all 7 steps in order |
| Error messages with no guidance | Every automated decision explained in plain-language output |

---

## Installation

```r
# install.packages("devtools")
devtools::install_github("EasternTechFusion/atspR")
```

---

## Pipeline

```
Raw data
   │
   ├─ (Stage 1) combine_datetime()      Merge date + time columns → POSIXct (optional)
   ├─ (Stage 2) fill_time_gaps()        Insert placeholder rows for missing timestamps
   │
   └─ ts_preprocess()                   ────────────────── 8 steps ──────────────────
   │    (Stage 3) standardize_na()      Convert sentinel values to NA
   │    (Stage 4) coerce_numeric()      Parse character columns to numeric
   │    (Stage 5) split_data()          Temporal train/test split
   │    (Stage 6) missing_analysis()    Summarise NAs; decide DROP or IMPUTE
   │    (Stage 7) handle_missing()      Drop rows or interpolate (linear/KNN)
   │    (Stage 8) visualize_data()      Scatter plots per variable vs time
   │    (Stage 9) scale_data()          MinMax / Z-score / Robust (auto)
   │    (Stage 10) cross_validate()     Walk-forward k-fold CV (optional)
   │
   └─ ts_export()                       Write all outputs to CSV
```

---

## Case 1: Date + Time in separate columns

**Sample data**

| Date | Time | Temp | Humid | Solar | WindDirection | WindSpeed | RainFall | VPD |
|---|---|---|---|---|---|---|---|---|
| 2024-01-01 | 08:00 | 25.12 | 80.04 | 14.0591 | 94.8218 | 0.0000 | 0.0000 | 0.2758 |
| 2024-01-01 | 09:00 | NA | 82.38 | 18.3274 | 91.2340 | 0.5200 | 0.0000 | 0.3124 |
| 2024-01-01 | 11:00 | 26.26 | 85.03 | 22.7810 | 88.6500 | 1.1400 | 0.0000 | 0.3892 |
| 2024-01-01 | 12:00 | 27.07 | 83.57 | NA | 90.1200 | 1.3300 | 0.1200 | 0.4015 |

> Hour 10:00 is missing entirely — `fill_time_gaps()` will insert it as a row of `NA`.

```{r case1, eval=FALSE}
library(atspR)

df <- combine_datetime(my_data,
                       date_col  = "Date",
                       time_col  = "Time",
                       new_col   = "datetime",
                       time_type = "string")

gap <- fill_time_gaps(df,
                      time_col = "datetime",
                      n        = 1,
                      unit     = "hour")

result <- ts_preprocess(data        = gap$data,
                        train_ratio = 0.8,
                        target_col  = "VPD",
                        k_folds     = 5L)
```

---

## Case 2: Datetime in one column

**Sample data**

| DateTime | Temp | Humid | Solar | WindDirection | WindSpeed | RainFall | VPD |
|---|---|---|---|---|---|---|---|
| 2024-01-01 08:00 | 25.12 | 80.04 | 14.0591 | 94.8218 | 0.0000 | 0.0000 | 0.2758 |
| 2024-01-01 09:00 | NA | 82.38 | 18.3274 | 91.2340 | 0.5200 | 0.0000 | 0.3124 |
| 2024-01-01 11:00 | 26.26 | 85.03 | 22.7810 | 88.6500 | 1.1400 | 0.0000 | 0.3892 |
| 2024-01-01 12:00 | 27.07 | 83.57 | NA | 90.1200 | 1.3300 | 0.1200 | 0.4015 |

> Hour 10:00 is missing entirely — `fill_time_gaps()` will insert it as a row of `NA`.

```{r case2, eval=FALSE}
my_data$datetime <- as.POSIXct(my_data$datetime,
                                format = "%Y-%m-%d %H:%M:%S")

gap <- fill_time_gaps(my_data,
                      time_col = "datetime",
                      n        = 1,
                      unit     = "hour")

result <- ts_preprocess(data        = gap$data,
                        train_ratio = 0.8,
                        target_col  = "VPD",
                        k_folds     = 5L)
```

---

## Case 3: Date only (daily data)

**Sample data**

| date | Temp | Humid | Solar | WindDirection | WindSpeed | RainFall | VPD |
|---|---|---|---|---|---|---|---|
| 2024-01-01 | 28.38 | 75.04 | 14.0591 | 94.8218 | 0.0000 | 0.0000 | 0.275835 |
| 2024-01-02 | 29.16 | 72.05 | NA | 91.2340 | 0.5200 | 0.0000 | 0.312410 |
| 2024-01-04 | 27.89 | 78.03 | 22.7810 | 88.6500 | 1.1400 | 2.4000 | 0.389200 |
| 2024-01-05 | 26.50 | 80.54 | 19.4320 | NA | 0.8800 | 0.0000 | 0.301770 |

> 2024-01-03 is missing entirely — `fill_time_gaps()` will insert it as a row of `NA`.

```{r case3, eval=FALSE}
gap <- fill_time_gaps(my_data,
                      time_col = "date",
                      n        = 1,
                      unit     = "day")

result <- ts_preprocess(data        = gap$data,
                        train_ratio = 0.8,
                        target_col  = "VPD",
                        k_folds     = 5L)
```

---

## Export Results

```{r export, eval=FALSE}
ts_export(result,
          dir          = "output",
          prefix       = "data",
          export_clean  = TRUE,
          export_report = FALSE)
# Saves:
#   output/data_train_scaled.csv
#   output/data_test_scaled.csv
#   output/data_data_clean.csv
```

---

## Cross-Validation

`atspR` uses **walk-forward validation** — the correct approach for
time-series data. Each validation fold is always preceded only by past
data, so no future information leaks into training.

```
Seed(20%)  Fold1  Fold2  Fold3  Fold4  Fold5
[─────────][─────][─────][─────][─────][─────]

Fold 1: train = seed              → val = fold1
Fold 2: train = seed + fold1      → val = fold2
Fold 3: train = seed + fold1+2    → val = fold3
...
```

```{r cv_example, eval=FALSE}
data(airquality)
clean <- airquality[complete.cases(airquality), ]
sp    <- split_data(clean, verbose = FALSE)
sc    <- scale_data(sp, method = "auto", verbose = FALSE)

cv <- cross_validate(
  sc,
  target_col     = "Ozone",
  k              = 5L,
  min_train_size = 0.2
)

cv$summary
cv$fold_results
```

---

## Available Functions

| Function | Description |
|---|---|
| `standardize_na()` | Convert custom missing indicators to `NA` |
| `coerce_numeric()` | Auto-convert character columns to numeric |
| `combine_datetime()` | Merge date + time columns → `POSIXct` |
| `fill_time_gaps()` | Insert missing timestamp rows |
| `missing_analysis()` | Analyse & visualise missing values |
| `handle_missing()` | Drop rows or impute (linear / KNN) |
| `visualize_data()` | Scatter plots per variable |
| `split_data()` | Temporal train / test split |
| `scale_data()` | Feature scaling (auto / minmax / zscore / robust) |
| `inverse_scale()` | Reverse scaling to original units |
| `cross_validate()` | Walk-forward k-fold CV |
| `ts_preprocess()` | Run all steps in one call |
| `ts_export()` | Export results to CSV |

---

## Citation

```
Sueppong Mueanchamnong and Pattharaporn Thongnim (2026). atspR: Automated
Time-Series Preprocessing in R. R package version 1.0.0.
https://github.com/EasternTechFusion/atspR
```
