---
title: "Getting Started with atspR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with atspR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4
)
library(atspR)
```

## Overview

`atspR` (**Automated Time-Series Preprocessing in R**) provides a complete,
modular preprocessing framework for time-series data — from raw input to
cross-validated, model-ready datasets.

---

## Installation

```r
devtools::install_github("EasternTechFusion/atspR")
```

---

## Case 1: Date + Time in separate columns

`combine_datetime()` รองรับ `time_type` 3 แบบ:

| `time_type` | ตัวอย่าง | คำอธิบาย |
|---|---|---|
| `"hour"` | `8`, `14` | Integer 0–23 |
| `"hhmm"` | `830`, `1430` | Integer HHMM |
| `"string"` | `"08:30"` | Character HH:MM |

```{r case1, eval=FALSE}
library(atspR)

# combine date + time columns → POSIXct
df <- combine_datetime(my_data,
                       date_col  = "Date",
                       time_col  = "Time",
                       new_col   = "datetime",
                       time_type = "string")

# fill missing timestamp rows (e.g. hour 10:00 absent)
gap <- fill_time_gaps(df,
                      time_col = "datetime",
                      n        = 1,
                      unit     = "hour")

result <- ts_preprocess(
  data        = gap$data,
  train_ratio = 0.8,
  target_col  = "VPD",
  k_folds     = 5L
)
```

---

## Case 2: Datetime in one column

```{r case2, eval=FALSE}
df_datetime <- data.frame(
  datetime = as.POSIXct(
    c("2024-01-01 08:00:00", "2024-01-01 09:00:00",
      "2024-01-01 11:00:00", "2024-01-01 12:00:00"),
    format = "%Y-%m-%d %H:%M:%S"
  ),
  Temp  = c(25.12, NA,    26.26, 27.07),
  Humid = c(80.04, 82.38, 85.03, 83.57),
  VPD   = c(0.276, 0.312, 0.389, 0.402)
)

# fill missing timestamp rows (e.g. hour 10:00 absent)
gap <- fill_time_gaps(df_datetime,
                      time_col = "datetime",
                      n        = 1,
                      unit     = "hour")

result <- ts_preprocess(
  data        = gap$data,
  train_ratio = 0.8,
  target_col  = "VPD",
  k_folds     = 5L
)
```

---

## Case 3: Date only (daily data)

```{r case3, eval=FALSE}
df_daily <- data.frame(
  date = as.Date(c("2024-01-01", "2024-01-02", "2024-01-04", "2024-01-05")),
  Temp  = c(28.38, 29.16, 27.89, 26.50),
  Humid = c(75.04, 72.05, 78.03, 80.54),
  VPD   = c(0.276, 0.312, 0.389, 0.302)
)

# fill missing timestamp rows (e.g. 2024-01-03 absent)
gap <- fill_time_gaps(df_daily,
                      time_col = "date",
                      n        = 1,
                      unit     = "day")

result <- ts_preprocess(
  data        = gap$data,
  train_ratio = 0.8,
  target_col  = "VPD",
  k_folds     = 5L
)
```

---

## Walk-Forward Cross-Validation

`atspR` uses **walk-forward validation** — the correct approach for
time-series. Each fold's validation set always follows its training set
in time.

```
Seed(20%)  Fold1  Fold2  Fold3  Fold4  Fold5
[─────────][─────][─────][─────][─────][─────]

Fold 1: train = seed              → val = fold1
Fold 2: train = seed + fold1      → val = fold2
Fold 3: train = seed + fold1+2    → val = fold3
```

```{r cv_example, eval=FALSE}
data(airquality)
clean <- airquality[complete.cases(airquality), ]
sp    <- split_data(clean, verbose = FALSE)
sc    <- scale_data(sp, method = "auto", verbose = FALSE)

cv <- cross_validate(
  sc,
  target_col     = "Ozone",
  k              = 5L,
  min_train_size = 0.2
)

cv$summary
cv$fold_results
```

---

## Export Results

```{r export, eval=FALSE}
ts_export(result, dir = "output", prefix = "data")
# Saves:
#   output/data_train_scaled.csv
#   output/data_test_scaled.csv
#   output/data_data_clean.csv
```

---

## Available Functions

| Function | Description |
|---|---|
| `standardize_na()` | Convert custom missing indicators to `NA` |
| `coerce_numeric()` | Auto-convert character columns to numeric |
| `combine_datetime()` | Merge date + time columns → `POSIXct` |
| `fill_time_gaps()` | Insert missing timestamp rows |
| `missing_analysis()` | Analyse & visualise missing values |
| `handle_missing()` | Drop rows or impute (linear / KNN) |
| `visualize_data()` | Scatter plots per variable |
| `split_data()` | Temporal train / test split |
| `scale_data()` | Feature scaling (auto / minmax / zscore / robust) |
| `inverse_scale()` | Reverse scaling to original units |
| `cross_validate()` | Walk-forward k-fold CV |
| `ts_preprocess()` | Run all steps in one call |
| `ts_export()` | Export results to CSV |
